---
title: "Grouped-variables"
author: "Lodewic van Twillert"
date: "10/24/2018"
output: 
  html_document:
    fig_caption: yes
    theme: cerulean
    toc_float: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r libraries}
# Load libraries
library(tidyverse)
library(broom)
library(ggthemes)
library(DT)
library(knitr)
```

```{r echo=F}
# Read the raw data
data <- read.csv("../data/raw/water_pump_set.csv")
labels <- read.csv("../data/raw/water_pump_labels.csv")
features <- read.csv("../data/raw/water_pump_features.csv")

# Merge labels and features into single dataset by id
df <- merge(data, labels, by="id")

# Let's add some more info to the feature info using tidyverse piping
feature_desc <- features %>% filter(Feature %in% colnames(data)) %>%
  mutate(example = sapply(data[1, -1], as.character), # First row of the data without the 'id' column
       unique = map_int(data[,-1],function(x) length(unique(x))),
       type = map_chr(data[,-1], function(x) class(unlist(x))), # 
       nonzero = map_int(data[,-1], function(x) sum(as.character(x) != "0")), # Nonzero values, assuming 0 usually means NA
       max_single = map_int(data[,-1], function(x) max(table(x))),      # Observations of most occuring value
       max_occurence = map_chr(data[,-1], function(x) names(which.max(table(x))))) # Highest occuring value
```

# Selecting the right features

Selecting the right features is extremely important! Especially in this case where we are dealing with
a majority of categorical features. Many of these categorical variables are redundant,
so let's first reduce the dimensionality of the data by selecting the most valuable and unique
variables.

## Handling feature groups 

The variables in the data are far from independent, some variables practically mean the same thing.
Some feature groups are so painfully obvious that we should clean these manually before moving on to
more generic analyses.

In the `Description` column, we can see some overlap in values. These features are likely to represent the same
variable, so we can reduce the dimensionality of the dataset just by smartly combining some variables.

Let's start with common sense and group some of the variables manually to prepare the data
for more automated pipelines. Here are some logical groups we can check,

  - scheme_name, scheme_management, installer, funder
  - extraction_type, extraction_type_group, extraction_type_class
  - management, management_group
  - scheme_management, scheme_name
  - payment, payment_type
  - water_quality, quality_group
  - quantity_quantity_group
  - source, source_class
  - waterpoint_type, waterpoint_type_group
  
### Scheme_name, scheme_management, installer, funder
 
These 4 features have far too many unique values to be used as categorical labels.
In some cases we can bin groups by grouping the least occurring data as 'other', but 
in these cases the arity is too high.

Together with spelling errors in these labels we won't spend the effort to transform
these values into something more useful with text processing.
 
```{r scheme management table}
data %>% 
    select(scheme_name, scheme_management, installer, funder) %>%
    mutate_all(as.character) %>%
    mutate_all(tolower) %>% # To lowercase since there are many typing 'errors' or inconsistencies
    group_by(scheme_name, scheme_management, installer, funder) %>%
    tally() %>% 
    arrange(desc(n)) %>% 
    head(10) %>%
    datatable()
```

The scheme names are a mess. There are non-ambiguous names, spelling errors,
numeric values, etc. Values such as like 'Adra', 'ADRA', 'WOULD BANK', etc.. are
variations on each other. We can reduce this already by using the lowercase version, 
but not enough to warrant the inclusion of this variable.

```{r scheme name table}
# Scheme names
data %>% select(scheme_name) %>% table() %>% sort(decreasing = T) %>% head()
# Half the data is missing data, and the rest is extremely scattered
# Furthermore there are many unambiguous names and spelling errors like 'Adra', 'ADRA', 'WOULD BANK', etc..
df %>%
    filter(scheme_name != "") %>%
    group_by(scheme_name, status_group) %>%
    tally() %>% head(20) %>%
    datatable()
# Based on this we remove the scheme_name entirely
```

```{r}
# Scheme management
data %>% select(scheme_management) %>% table() %>% sort(decreasing = T) %>% 
  head()

# Plot label per scheme_management
df %>% select(scheme_management, status_group) %>%
    ggplot(aes(x = scheme_management, fill = status_group)) +
    geom_bar() + coord_flip() +
    ggtitle("Water pomp status voor management_scheme") +
    theme_fivethirtyeight()

# Most of the water pumps are from a handful of installers. 
# We could consider grouping all the low occurences into a group 'other'
# Plot label per scheme_management
scheme_management_groups = df %>% select(scheme_management, status_group) %>%
    group_by(scheme_management) %>%
    tally() %>%
    top_n(10, n)

cat(sprintf("Scheme management in 10 groups accounts for %1.2f%% of the data. And just VWC is %1.2f%%", 
            100 * sum(scheme_management_groups$n) / nrow(df),
            100 * max(scheme_management_groups$n) / nrow(df)))
```

#### Installer and funder

The installer and funder values have many levels. Again a lot of data-entry errors
seem apparent in the data, and when transformed to lower-case the installer and funder
variables overlap for over 50%. Because of the arity of this data we remove both.

  - Possible to do string matching

```{r}
# installer
data %>% select(installer) %>% table() %>% sort(decreasing = T) %>% head()
length(unique(data$installer))

# Most of the water pumps are from a handful of installers. 
# We could consider grouping all the low occurences into a group 'other'
# Plot label per scheme_management
installer_groups = df %>% select(installer, status_group) %>%
    group_by(installer) %>%
    tally() %>%
    top_n(20, n)

# Plot installer versus status_group to verify 
df %>% select(installer, status_group) %>%
    mutate(installer = ifelse(installer %in% installer_groups$installer, as.character(installer), "other"),
           installer = ifelse(installer %in% c("", "0"), "unknown", installer)) %>%
    ggplot(aes(x = installer, fill = status_group)) +
    geom_bar() + coord_flip() +
    ggtitle("Water pomp status voor installer", subtitle = "Kleine groepen samengevoegd in 'other'") +
    theme_fivethirtyeight() 

# Show counts of installer labes, only top20
# Notice that many of these have about the same number of observations and grouping all the others into
# 'other' is quite arbitrary since the 'other' group will still be half the data. 
# The number of 'larger' groups and the number of spelling errors in the data require more detailed feature engineering
# like some kind of string matching- which we will not spend our time on.
# THIS FEATURE WILL BE REMOVED FROM THE MODELLING DATA
```

```{r}
# funder
data %>% select(funder) %>% table() %>% sort(decreasing = T) %>% head()

cat(sprintf("Of the %i unique funder names, %i overlap with the installer names!",
            length(unique(data$funder)),
            length(intersect(tolower(as.character(data$funder)), tolower(as.character(data$installer))))))
# Due to the high overlap between funder and installer, we can assume that these two are heavily related and 
# will probably give us the same difficulty as the funder names by themselves.
# WE REMOVE THIS FEATURE FROM THE DATASET!
```




